---
title: "Sample size calculation for survival outcomes"
author: "Yongxi Long"
format: 
  html:
    self-contained: true
editor: visual
---

## Aim

Figure out how to better explain sample size calculation for survival outcomes

[Online calculator](https://sample-size.net/sample-size-survival-analysis/)

The reference is from [This paper](https://www.jstor.org/stable/2531021)

## Two-step calculation

### Determine the number of events needed

"Since the usual tests of treatment effect, when covariates are adjusted for, are asymptotically equivalent, I shall derive a sample-size formula for the score test."

Let:

-   $\alpha,\beta$ be the type I error and type II error respectively
-   $\Delta$ be the hazard ratio
-   $P_0,P_1$ be the randomization ratio
-   $ED$ be the expected number of deaths on the trial

The score test statistic is asymptotically normal with unit variance and mean $log(\Delta)\sqrt{(P_0 P_1 ED)}$

By solving $Power=P(S>z_{1-\alpha/2})+P(S<z_\alpha/2)=1-\beta$, we get:

$ED = \frac{(z_{1-\alpha/2}+z_{1-\beta})^2}{P_0P_1log^2(\Delta)}$

```{r,echo=FALSE}
# function to calculate number of events needed
events.needed <- function(alpha=0.05,beta=0.2,HR=0.8,P0=0.5,P1=0.5)
{
  (qnorm(1-alpha/2)+qnorm(1-beta))^2/(P0*P1*log(HR)^2)
}
```

```{r}
# set type I and type II error
alpha <- 0.05; beta <- 0.2
# give expected treatment effect in terms of hazard ratio
HR <- 0.6
# set the randomization ratio
P0 <- 0.5
P1 <- 1-P0
# get the number of events needed
events <- events.needed(alpha,beta,HR,P0,P1)
events
```

### Determine the expected proportion of patients having the events

-   We divided the whole study duration into two parts:

    1.  an accrual period, a, the period during which patients enter the study

    2.  and a follow-up period, f, the period from the end of accrual until the analysis of the data.

    The range of follow-up time for patients is from f to f+a, assuming no loss-to-follow-up

-   The approximate proportion of patients having the events depends on the survival curve, the length of the accrual time and the follow-up period

```{r}
# function to approximate prop of having the event from exponential survival
# t is the median survival
# a the accural period
# f the follow-up period
prop.event <- function(t,f,a)
{
  1-exp(-0.69*f/t)*(1-exp(-0.69*a/t))/(0.69*a/t)
}
```

## Now we assume we have the total study duration fixed at 4 years

Let's see the impact of the accrual period

```{r}
# set the hazard for group 0
total.duration <- 4
hazard0 <- 0.5
median.surv0 <- log(2)/hazard0
hazard1 <- hazard0*HR
median.surv1 <- log(2)/hazard1
```

### If patients appear all at once at the beginning (a=0)

```{r}
# can't give a cuz the denominator would be zero
a = 0.01 ;f=total.duration -a
# proportion of patients having the event in group 0
prop.event0 <- prop.event(t=median.surv0,f=f,a=a)
# proportion of patients having the event in group 1
prop.event1 <- prop.event(t=median.surv1,f=f,a=a)
# proportion of patients having the event combined
prop.event.comb <- P0*prop.event0+P1*prop.event1
# total number of patients needed
events/prop.event.comb
```

### If spend 1/4 of the time on recruiting

This means the average follow-up time for all patients is a/2 + f = 3.5, assuming constant accrual rate

```{r,echo=FALSE}
# can't give a cuz the denominator would be zero
a = 1 ;f=total.duration -a
# proportion of patients having the event in group 0
prop.event0 <- prop.event(t=median.surv0,f=f,a=a)
# proportion of patients having the event in group 1
prop.event1 <- prop.event(t=median.surv1,f=f,a=a)
# proportion of patients having the event combined
prop.event.comb <- P0*prop.event0+P1*prop.event1
# total number of patients needed
events/prop.event.comb
```

### If spend 1/2 of the time on recruiting

This means the average follow-up time for all patients is a/2 + f = 3, assuming constant accrual rate

```{r,echo=FALSE}
# can't give a cuz the denominator would be zero
a = 2 ;f=total.duration -a
# proportion of patients having the event in group 0
prop.event0 <- prop.event(t=median.surv0,f=f,a=a)
# proportion of patients having the event in group 1
prop.event1 <- prop.event(t=median.surv1,f=f,a=a)
# proportion of patients having the event combined
prop.event.comb <- P0*prop.event0+P1*prop.event1
# total number of patients needed
events/prop.event.comb
```

### If spend all of the time on recruiting

This means the average follow-up time for all patients is a/2 + f = 2, assuming constant accrual rate

```{r,echo=FALSE}
# can't give a cuz the denominator would be zero
a = 4 ;f=total.duration -a
# proportion of patients having the event in group 0
prop.event0 <- prop.event(t=median.surv0,f=f,a=a)
# proportion of patients having the event in group 1
prop.event1 <- prop.event(t=median.surv1,f=f,a=a)
# proportion of patients having the event combined
prop.event.comb <- P0*prop.event0+P1*prop.event1
# total number of patients needed
events/prop.event.comb
```
